MSB1015 Project

### 2024-10-10 Code/MSB1015-Project.Rmd
Update MSB1015-Project.Rmd
Fixed some library issues, and divided the random forest chunk into multiple sub-chunks for easier use. Additionally, I added code to perform unsupervised random forest and visualised it using a function I made.

data <- read_csv("METABRIC_RNA_Mutation.csv")
```

@@

# Define thresholds for filtering
lower_threshold <- mean_score * 0.95 # might do 5%
upper_threshold <- mean_score * 1.05 # only low visualise with PCA

# Filter out rows where the score is outside of 3 standard deviations from the mean
data_filtered <- data_filtered %>%
  filter(pred >= lower_threshold) %>%
  filter(pred <= upper_threshold) %>%
  select(-pred)

# Check for inconsistencies
result <- lapply(data_filtered, function(i) {
  if (is.factor(i)) {
    unique(i)
  }
})
result <- Filter(Negate(is.null), result)
# Print the result
print(result)
```

# 2. The random forest
```{r random forest, echo=FALSE}
# Load required packages
library(randomForest)
library(caTools)
library(caret)
library(nestedcv)
library(mlbench)
library(pbapply)
library(ROCR)
library(PRROC)
library(pROC)

categorise_prognosis <- function(months) {
  case_when(
    months <= 60 + (60 * 0.25) ~ "will_die_within_5_years",
    months > 60 + (60 * 0.25) ~ "will_die_within_10_years",
  )
}

# Process the data
categorical_data <- data_filtered %>%
  select(age_at_diagnosis:death_from_cancer) %>%
  filter(death_from_cancer != "Died of Other Causes") %>%
  filter(overall_survival == 0) %>%
  mutate(
    prognosis = categorise_prognosis(overall_survival_months)
  ) %>%
  select(-death_from_cancer, -overall_survival, -overall_survival_months)

genetic_data <- data_filtered %>%
  select(brca1:stmn2_mut, overall_survival_months, death_from_cancer, overall_survival) %>%
  filter(death_from_cancer != "Died of Other Causes") %>%
  filter(overall_survival == 0) %>%
  mutate(
    prognosis = categorise_prognosis(overall_survival_months)
  ) %>%
  select(-death_from_cancer, -overall_survival, -overall_survival_months)

full_data <- data_filtered %>%
  filter(death_from_cancer != "Died of Other Causes") %>%
  filter(overall_survival == 0) %>%
  mutate(
    prognosis = categorise_prognosis(overall_survival_months)
  ) %>%
  select(-death_from_cancer, -overall_survival, -overall_survival_months)

# Split the data into training and testing sets
set.seed(111, "L'Ecuyer-CMRG")

trainIndex <- createDataPartition(categorical_data$prognosis, p = .7, 
                                  list = FALSE, 
                                  times = 1)

train <- categorical_data[ trainIndex,]
test  <- categorical_data[-trainIndex,]

@@

tuneGrid <- expand.grid(mtry = 2:8)
ntrees <- c(1000, 2000) 
nodesize <- c(1, 5)

# List for the combinations
list <- 4

@@

# Initialize an empty list to store the metrics
metrics_list <- list()

# Loop over the indices in the provided list
for (i in list) {
  # Apply the metrics function to each model in store_models
  metrics_list[[names(store_models)[i]]] <- metrics(store_models[[i]])
}

@@

# Get an overview of the best model
summary(best_model)

# Calculate accuracy on the train set for the best model
# model_predict <- best_model$final_fit$finalModel$predicted
# pred_table_training <- table(model_predict, y_train)
# confusionMatrix(pred_table_training)
pred_table_training <- table(best_model$output$predy, best_model$output$testy)
train_confusion <- confusionMatrix(pred_table_training)
train_confusion

# Call the plotting function for the confusion matrix
source("~/University/MSB/MSB1015/confusionPlot.R")
confusionPlot(train_confusion, title = "Confusion Matrix (Training Set)")

@@

@@

# Plot OOB
ggplot(err_rate, aes(x = ntrees, y = OOB)) +
  geom_line(color = "skyblue1", size = 1) +
  labs(x = "Number of Trees", y = "OBB Error", title = "Out Of Bags (OOB) error") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, size = 14, face = "bold"))

###############
### TESTING ###
###############

@@

confusionPlot(test_confusion, title = "Confusion Matrix (Testing Set)")

>>>>

# data <- read_csv("METABRIC_RNA_Mutation.csv")
```

@@

# Store the original number of rows before filtering
original_count <- nrow(data_filtered)

lower_threshold <- mean_score - 2*sd_score

# Filter out rows where the score is outside of 2 standard deviations from the mean
data_filtered <- data_filtered %>%
  filter(pred >= lower_threshold) %>%
  select(-pred)  # Remove the 'pred' column

# Store the number of rows after filtering
filtered_count <- nrow(data_filtered)

# Calculate the number of eliminated rows
eliminated_count <- original_count - filtered_count

# Print the number of eliminated variables
cat("Number of eliminated variables:", eliminated_count, "\n")
```

# 2. The random forest
```{r data tranformation for random forest, echo=FALSE}

# Load required packages
library(randomForest)
library(caTools)
library(caret)
library(nestedcv)
library(mlbench)
library(pbapply)
library(ROCR)
library(PRROC)
library(pROC)
library(gridExtra)

# Create a function to create the prognosis variable
categorise_prognosis <- function(months) {
  case_when(
    months <= 60 + (60 * 0.25) ~ "will_die_within_5_years",
    months > 60 + (60 * 0.25) ~ "will_die_within_10_years",
  )
}

# Process the data
categorical_data <- data_filtered %>%
  select(age_at_diagnosis:death_from_cancer) %>%
  filter(death_from_cancer != "Died of Other Causes") %>%
  filter(overall_survival == 0) %>%
  mutate(
    prognosis = as.factor(categorise_prognosis(overall_survival_months))
  ) %>%
  select(-death_from_cancer, -overall_survival, -overall_survival_months) %>%
  as.data.frame()

source("~/University/MSB/MSB1015/funPCA.R") # Change the directory as needed

# Call the function and plot
pca_plots_categorical <- pca_proximity_plot(categorical_data, "prognosis")
grid.arrange(grobs = pca_plots_categorical, ncol = 2)

genetic_data <- data_filtered %>%
  select(brca1:stmn2_mut, overall_survival_months, death_from_cancer, overall_survival) %>%
  filter(death_from_cancer != "Died of Other Causes") %>%
  filter(overall_survival == 0) %>%
  mutate(
    prognosis = as.factor(categorise_prognosis(overall_survival_months))
  ) %>%
  select(-death_from_cancer, -overall_survival, -overall_survival_months) %>%
  as.data.frame()

# Call the function and plot
pca_plots_genetic <- pca_proximity_plot(genetic_data, "prognosis")
grid.arrange(grobs = pca_plots_genetic, ncol = 2)

complete_data <- data_filtered %>%
  filter(death_from_cancer != "Died of Other Causes") %>%
  filter(overall_survival == 0) %>%
  mutate(
    prognosis = as.factor(categorise_prognosis(overall_survival_months))
  ) %>%
  select(-death_from_cancer, -overall_survival, -overall_survival_months) %>%
  as.data.frame()

# Call the function and plot
pca_plots_complete <- pca_proximity_plot(complete_data, "prognosis")
grid.arrange(grobs = pca_plots_complete, ncol = 2)
```

# 2.1 The random forest
```{r random forest, echo=FALSE}

# Set the seed for reproducibility of random results using the L'Ecuyer-CMRG method
set.seed(111, "L'Ecuyer-CMRG")

# Create a partition to split the data into training and testing sets
# The target variable for stratification is 'prognosis', with 70% of the data for training
trainIndex <- createDataPartition(categorical_data$prognosis, p = .7, 
                                  list = FALSE,  # Return indices as a matrix
                                  times = 1)     # Perform the partitioning once

# Subset the original data to create the training set using the indices from trainIndex
train <- categorical_data[trainIndex,]

# Subset the original data to create the testing set using the remaining indices
test  <- categorical_data[-trainIndex,]

@@

tuneGrid <- expand.grid(mtry = 2:8)

# The model with the highest AUC that I found was with 2000 trees and node size of 1. 
# When creating the model these values can be changed.
ntrees <- c(2000) 
nodesize <- c(1)

# List for the combinations: should have all the possible combinations between ntrees and nodesize.
list <- 1

@@

# 2.2 Finding the best model
```{r best model, echo=FALSE}
# Initialize an empty list to store the metrics
metrics_list <- list()

# Loop over the indices in the provided list
for (i in seq_along(store_models)) {
  # Apply the metrics function to each model in store_models
  metrics_list[[names(store_models)[i]]] <- metrics(store_models[[i]])
}

@@

# Get an overview of the best model
summary(best_model)
```

# 2.3 ROC and PR curves on the best model
```{r accuracy and plotting, echo=FALSE}
# Calculate accuracy on the train set for the best model
pred_table_training <- table(best_model$output$predy, best_model$output$testy)
train_confusion <- confusionMatrix(pred_table_training)
train_confusion

# Call the plotting function for the confusion matrix
source("~/University/MSB/MSB1015/confusionPlot.R") # Change the directory as needed
confusionPlot(train_confusion, title = "Confusion Matrix (Training Set)")

@@

importance_sorted <- importance[order(importance$will_die_within_10_years, decreasing = TRUE), ]
head(importance_sorted)

@@

# Plot OOB
ggplot(err_rate, aes(x = ntrees, y = OOB)) +
  geom_line(color = "skyblue1", linewidth = 1) +
  labs(x = "Number of Trees", y = "OBB Error", title = "Out Of Bags (OOB) error") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, size = 14, face = "bold"))
```

# 2.4 Testing
```{r testing, echo=FALSE}

@@

confusionPlot(test_confusion, title = "Confusion Matrix (Testing Set)")
```
